Jug: Software for Reproducible Computation In Python

### 2016-11-12 paper.tex
ENH Update paper
Back several assertions with appropriate cites
Improve 'related tools' section

\hypersetup{%
    bookmarks=true,
    colorlinks=true,
    pdfauthor={Luis Pedro Coelho},
    citecolor=pondgreen,
    urlcolor=toastedchilipowder,
    linkcolor=toastedchilipowder,
}
\let\code\texttt
\addbibresource{references.bib}

@@

In addition to being valuable to the wider community, which may wish to
reproduce or extend a published analysis, reproducible research practices allow
for better control over the project by the authors themselves. For example, if
necessary parameters to run a pipeline are kept separately from the code that
implements it, this leads to error-prone analysis and opens up the possibility
that when the results are to be written up for publication, the researcher will
no longer be able to even completely describe the process that led to them.

For large projects, the use of multiple computational ccores (either in the
same machine or distributed across a cluster) is necessary to obtain results in
a useful time frame. Furthermore, it is often the case that, as the project
evolves, it becomes necessary to save intermediate results while down-stream
analyses are designed (or redesigned) and implemented. Under many frameworks,
this causes having a single point of entry (often in the form of a single
script or a single main programme) for the computation becomes increasingly
difficult.

@@

\section{Introduction}
The value of reproducible research in computational fields has been recognized
in several areas \citep{Vandewalle2009,Nordlie2009}. Although some of the
reports are still more forward looking rather than descriptive of current
practice, several implementations of reproducible papers (or executable papers)
have been proposed towards the goal of reproducing published analyses.

These solutions do not necessarily scale to large problems, those that take
several days, months, or years of CPU time. For very large problems,
specialized solutions are needed~\citep{mapReduce}. There is a range of
medium-sized problems that can be successfully tackled on a computer cluster
with a few hundred nodes or less (so that a computation that would take a
CPU-year can produce results in less than a day).

The typical ad hoc solution to this problem is to save intermediate files on
disk. Often the design of the computation itself takes several iterations as
intermediate steps are improved. Thus, some intermediate results need to be
recomputed. This involves a great amount of human management of the state of
the computation, breaking it up into pieces, and issuing jobs on the cluster.

@@

However, unlike make, jug is written in Python and tasks can consist of any
Python function and its arguments. This can include running external commands
and calling routines written in other languages as Python has many tools to
interface with the rest of the system.

A jugfile (the file which specifies which tasks need to be run---named by
analogy to the makefiles of make) is a simple Python script with some added
notations (we will see the only a few, small, changes are needed to transform a
conventional Python script into a Jugfile).

@@

The filesystem can support all of the above operations if the backend is coded
correctly to avoid race conditions. This is the default backend, identified
simply by a directory name. Inside this directory, files named by a hexadecimal
representation of their hashes. Objects are saved using Python's \code{pickle}
module with zlib compression. As a special case, \code{numpy} arrays
\citep{numpy} are saved to disk directly. This special case was introduced as
\code{numpy} arrays are a very common data type in scientific programming and
saving them directly allows for very fast saving and loading (they are
represented on disk as a header followed by the binary information they
contain).

@@

The output of computation can be obtained from jug in three ways: (1) One can
simply write a task that writes the output to a file in the desired format. (2)
Outputs can be inspected interactively using the \emph{shell} subcommand. It is
expected that the first option will be used for the final version of the
computation, whilst the second one is most helpful during development. Finally,
(3) jug can be loaded as a library from a Python script and jug computation
outputs can serve as inputs for further computation.

The shell subcommand, invokes an IPython instance with all the objects in the
jugfile loaded. IPython is an enhanced interactive shell for Python
\citep{Perez2007}. A few functions are added to the namespace, in particular,
\code{value} can load the results of a task object if necessary.

@@

=========
Jug Shell
=========

@@

Enjoy...

@@

Figure~\ref{fig:jug-shell-interaction} shows a possible interaction session
with jug shell. While having to explicitly load all the results is a bit
bothersome, it can be better than pre-loading all of them. It is much faster at
startup. Consider too that the user might not load more than a few objects. In
some cases, loading all of the objects simultaneously might even be impossible
due to memory constraints. Further, this allows exploration of the task
structure for debugging.

This is a feature which can be used to debug one's own code or to explore the
results of a computation by others.

@@

If a developer improves an intermediate step in a pipeline (e.g., fixes a bug)
and wishes to obtain new results of the computation, then all outputs from that
step and downstream must be recomputed, but results from upstream and unrelated
processes can be reused. Formally, in the task DAG, affected tasks and their
descendants must be recomputed. Without tool support, this can be a very
error-prone operation: by not removing enough of the intermediate files, it was
easy to generate an irreproducible state where part of the computations were
made using one version of the code and part with another.

Therefore, jug adds support for result invalidation. When a results from a task
are invalidated, all tasks which, directly or indirectly, depend on them are
also marked for deletion.

It is still necessary to manually invalidate tasks. An alternative, would be to
take code which implements the task into account while computing the hash. This
would mark any results computed with this code as outdated if the code changed.
While this would add another layer of protection, it would still be possible to
make mistakes. If the function depended on other functions, especially if this
was done in a dynamic way, it could be hard to discover all dependencies (cf.\
sumatra~\footnote{Sumatra is a tool to collect all dependencies that are
necessary to run a particular computation. Its webpage is located at
\url{http://neuralensemble.org/trac/sumatra}.}). Additionally, even minute
refactoring of the code would lead to over eager recomputation. This could make
the developer wary of making improvements in their code, resulting in overall
worse code.

@@

\section{Methods}
\section{Example}

I will present an example from my research. Although it introduces some
unnecessary complexity, it has the advantage of being a real-life example of
the usage of jug.

The goal of this project was to demonstrate that using local features, in
particular speeded-up robust features (henceforth,
\textsc{surf})~\cite{surf-paper} improves over image-global features for
classification of microscope images. A variation of standard \textsc{surf},
\textsc{surf}-ref, was introduced and tested.

To evaluate classification, the dataset is broken up into 10~pieces and each of
the pieces is held-out of the analysis and then used to evaluate analysis (the
final result is the average of all ten values). This is known as
cross-validation. In this case, I use another library (which I also wrote)
which has a built-in method for jug-aware cross-validation.

\begin{python}
from jug import TaskGenerator, CachedFunction

# pyslic is used for subcellular location analysis
import pyslic

# milk (machine learning toolkit)
from milk.ext.jugparallel import nfoldcrossvalidation

datasets = [ 'data0', 'data1', ]
featuresets = [ 'field-dna+', 'surf', 'surf-ref', ]

def load(dset):
    '''
    This function assumes that the images are stored in
    ../data/dset with filenames matching the pattern
    label-(protein|dna)-([0-9]).tiff
    '''
    from os import listdir
    images = []
    base = '../data/'+dset
    for path in listdir(base):
        if not 'protein' in path: continue
        label = path.split('-')[0]
        # We only store paths and will load data on demand
        # This save memory.
        im = pyslic.Image({
                'protein': base + path,
                'dna' : base + path.replace('protein','dna')
                })
        im.label = label
        images.append(im)
    return images

# pyslic.computefeatures takes an image and an identifier of the
# feature set and returns a set of features (as a numeric array).
#
# TaskGenerator can be used directly to adapt existing functions
computefeatures = TaskGenerator(pyslic.computefeatures)

# or it can be used as a decorator on a newly defined function
@TaskGenerator
def output(cmatrix, dset, fset):
    output = file('results/cmatrix-'+dset+'-'+fset+'.txt', 'w')
    print >>output, cmatrix
    output.close()

features = []
for dset in datasets:
    images = CachedFunction(load, dset)
    labels = [im.label for im in images]
    for fset in featuresets:
        features = [computefeatures(im, fset) for im in images]
        cmatrix = nfoldcrossvalidation(features, labels)
        output(cmatrix, dset, fset)
\end{python}

@@

\section{Development of Jug}
Jug is available under the MIT software license, which grants the user right to
use, copy, and modify the code almost at will. It is developed using the git
version control tool, with the project being hosted on the github
platform.\footnote{The jug project is available at
\url{https://www.github.com/luispedro/jug}.} An open mailing-list provides
support and reported bugs are generally fixed within hours.

Jug includes a full test suite. There are no known bugs.

@@

\citet{Fomel2007} proposed a system built on top of Scons which shared
superficially similar design. Scons, like Make, is a build system. It supports
spawning parallel jobs using multiple threads. In the original use case, the
tasks are delegated to other commands and the operating system can take care of
parallelism. If, however, the tasks are computationally intensive in Python,
contention for the global interpreter lock will limit the amount of real
parallelism.\footnote{In the most commonly used Python interpreters, there is a
lock that prevents more than one thread from simultaneously executing
interpreted code, although they can execute non-interpreted code, as calling an
external programme or certain external libraries.} Ruffus is another
Python-based solution which supports parallel execution of pipelines
\citep{Goodstadt16092010}. There is a strong dependency on the filesystem,
which makes its use very different from jug usage, which strives to make its
usage similar to straight Python usage. Using the Ruby programming language,
\citet{mishima2011} proposed Pwrake, which supports parallel execution of
bioinformatics workflows written (or accessible) in that language.

Of interest, is also the \textsc{IncPy} implementation of a Python interpreter
with automatic persistent memoization~\cite{Guo_IncPy,Guo_IncPy_pre}. The
arguments made for that system apply to Jug as well, with a few differences.
Their system implements automatic memoization, while, using Jug, the user needs
to automatically annotate functions to memoize using \texttt{TaskGenerator}.
This extra control can be necessary to avoid a proliferation of intermediate
results when these are very large (often a very large intermediate output is
not necessary, and only a summary must be kept). Additionally, Jug supports
running tasks in parallel, a functionality that is absent from \textsc{IncPy}.

@@

Unlike most previous tools for reproducible research, jug focuses on the
development of the computation as much as its communication. In fact, when it
comes to communication other tools might be better suited as they combine
written exposition with computer code. They can be used in combination with jug
as they do not often provide caching and distribution, needed for large
projects. While the simplest mode of jug use is to have a single
\textit{jugfile}, a jugfile can import the results of another. For example, a
first jugfile might perform all heavy-duty computation and be run on a
computing cluster. A second script could be embedded inside an executable paper
to generate tables and plots \citep{Delescluse2011}. This faster script could
be run as part of building the paper (while the longer script would take too
long).

Jug improves the pipeline development experience and makes the programming
researcher more productive and less error-prone. As it is based on Python, a
widely used scientific programming language, it can be used in a known
environment without a long learning process.

Jug was not designed to compete with large scale frameworks such as Hadoop or
Spark, which scale better to very large projects. However, those frameworks
have much higher costs in terms of development time (code must be written
especially for the framework) and overhead. For scientists using small to
medium-sized cluster (hundreds, not thousands, of nodes), with a shared
file-system (as opposed to the distributed file-systems necessary for truly
large amounts of data) jug provides a better trade-off than those
higher-powered alternatives in a familiar programming language.

>>>>

\hypersetup{%
    bookmarks=true,
    colorlinks=true,
    pdfauthor={Luis Pedro Coelho},
    citecolor=pondgreen,
    urlcolor=toastedchilipowder,
    linkcolor=toastedchilipowder,
}
\let\code\texttt
\addbibresource{references.bib}
\addbibresource{readcube_export.bib}

@@

In addition to being valuable to the wider community, which may wish to
reproduce or extend a published analysis, reproducible research practices allow
for better control over the project by the original authors themselves. For
example, if necessary parameters to run a pipeline are kept separately from the
code that implements it, this leads to error-prone analysis and opens up the
possibility that, when the results are to be written up for publication, the
researcher will no longer be able to even completely describe the process that
led to them.

For large projects, the use of multiple computational cores (either in a
multi-core machine or distributed across a compute cluster) is necessary to
obtain results in a useful time frame. Furthermore, it is often the case that,
as the project evolves, it becomes necessary to save intermediate results while
down-stream analyses are designed (or re-designed) and implemented. Under many
frameworks, this causes having a single point of entry (often in the form of a
single script or a single main programme) for the computation becomes
increasingly difficult.

@@

\section{Introduction}
The value of reproducible research in computational fields has been recognized
in several areas
\citep{donoho2009reproducible,v2009reproducible,nordlie2009towards,fomel2015reproducible,leveque2012reproducible,dudley2010reproducible}.
Although some of the reports are still more forward looking rather than
descriptive of current practice, several implementations of reproducible papers
(or executable papers) have been proposed towards the goal of reproducing
published analyses.

These solutions do not necessarily scale to large problems, those that take
several days, months, or years of CPU time. For very large problems,
specialized solutions are needed~\citep{dean2008mapreduce,zaharia2010spark}.
There is a range of medium-sized problems that can be successfully tackled on a
computer cluster with a small number of nodes or even taking advantage of a
single multicore machine.

The typical ad hoc solution to this problem is to save intermediate files on
disk. A limitation of this approach is that often the design of the computation
itself takes several iterations as intermediate steps are improved. Thus, some
intermediate results need to be recomputed. This involves a large amount of
human management of the state of the computation, breaking it up into pieces,
and issuing jobs on the cluster.

@@

However, unlike make, jug is written in Python, a general purpose programming
language which is widely used in scientific programming \citep{prabhu2011a}.  A
Task in Jug can consist of any Python function and its arguments. The task can
include running external commands and calling routines written in other
languages as Python has many tools to interface with the rest of the system
\citep{beazley1996swig,beazley2003automated,behnel2011cython}.

A Jugfile (the file which specifies which tasks need to be run---named by
analogy to the Makefiles of make) is a simple Python script with some added
notations (we will see the only a few, small, changes are needed to transform a
conventional Python script into a Jugfile).

@@

The filesystem can support all of the above operations if the backend is coded
correctly to avoid race conditions. This is the default backend, identified
simply by a directory name. Inside this directory, files named by a hexadecimal
representation of their hashes. Objects are saved using Python's \code{pickle}
module with zlib compression. As a special case, \code{numpy} arrays
\citep{walt2011the} are saved to disk directly. This special case was
introduced as \code{numpy} arrays are a very common data type in scientific
programming and saving them directly allows for very fast saving and loading
(they are represented on disk as a header followed by the binary information
they contain).

@@

The output of computation can be obtained from jug in three ways: (1) One can
write a task that writes the output to a file in the desired format. (2)
Outputs can be inspected interactively using the \emph{shell} subcommand. It is
expected that the first option will be used for the final version of the
computation, whilst the second one is most helpful during development. Finally,
(3) jug can be loaded as a library from a Python script and jug computation
outputs can serve as inputs for further computation. This can be performed from
inside a Jupyter notebook \citep{kluyver2016jupyter}.

The shell subcommand, invokes an Jupyter instance with all the objects in the
jugfile loaded. The Jupyter console is an enhanced interactive shell for Python
\citep{perez2007ipython}. A few functions are added to the namespace, in
particular, \code{value} can load the results of a task object if necessary.

@@

@@

@@

Figure~\ref{fig:jug-shell-interaction} shows a possible interaction session
with jug shell. While having to explicitly load all the results may be
bothersome, it is both much faster at start up and the user might not load more
than a few objects throughout their session. In some cases, loading all of the
objects simultaneously might even be impossible due to memory constraints.
Furthermore, this allows exploration of the task structure for debugging.

@@

If a researcher improves an intermediate step in a pipeline (e.g., fixes a bug)
and wishes to obtain new results of the computation, then all outputs from that
step and downstream must be recomputed, but results from upstream and unrelated
processes can be reused. Formally, in the task DAG, affected tasks and their
descendants must be recomputed. Without tool support, this can be a very
error-prone operation: by not removing enough of the intermediate files, it was
easy to generate an irreproducible state where part of the computations were
made using one version of the code and part with another.

Therefore, jug adds support for result invalidation. When a results from a task
are invalidated, all tasks which (directly or indirectly) depend on them are
also invalidated.

It is still necessary to manually invalidate tasks. An alternative, would be to
take code which implements the task into account while computing the hash. This
would mark any results computed with this code as outdated if the code changed.
While this would add another layer of protection, it would still be possible to
make mistakes. If the function depended on other functions, especially if this
was done in a dynamic way, it could be hard to discover all dependencies (cf.\
sumatra, which does capture all dependencies that are necessary to run a
particular computation \citep{davison2012automated}). Additionally, even minute
refactoring of the code would lead to over eager recomputation. This could make
the developer wary of making improvements in their code, resulting in overall
worse code.

@@

\section{Example}
This section presents an edited version of the code used in a previously
published study of computer vision techniques for bioimage
analysis~\cite{coelho2013determining}. The goal of this project was to
demonstrate that using local features, in particular speeded-up robust features
(henceforth, \textsc{surf}) improves over image-global features for
classification of microscope images.

To evaluate classification, the dataset is broken up into 10~pieces and each of
the pieces is held-out of the analysis and then used to evaluate analysis (the
final result is the average of all ten values). This is known as
cross-validation.

\begin{python}
from jug import TaskGenerator, CachedFunction
import mahotas as mh
from sklearn.cross_validation import KFold

def load():
    '''
    This function assumes that the images are stored in
    data/<dset> with filenames matching the pattern
    label-protein-([0-9]).tiff
    '''
    from os import listdir
    images = []
    base = './data/'
    for path in listdir(base):
        if not 'protein' in path: continue
        label = path.split('-')[0]
        # We only store paths and will load data on demand
        # This save memory.
        im = (base + path, label)
        images.append(im)
    return images

# TaskGenerator can be used with any function, including builtins:
jug_sum = TaskGenerator(sum)

@TaskGenerator
def computefeatures(im):
    im = mh.imread(im)
    return mh.features.haralick(im)

@TaskGenerator
def output(cmatrix, dset, fset):
    with open('results/cmatrix-'+dset+'-'+fset+'.txt', 'w') as output:
        output.write(cmatrix)
        output.write('\n')

@TaskGenerator
def train_test1(features, labels, train, test):
    from sklearn import linear_model
    from sklearn import metrics
    clf = linear_model.LogisticRegression()
    clf.fit(features[train], labels[train])
    preds = clf.predict(features[test])
    return metrics.confusion_matrix(labels[test], preds)

images = CachedFunction(load)
features = []
images = []
for im,label in images:
    labels.append(label)
    features.append(haralick(im))
cmats = []
for train,test in KFold(len(features), n_folds=10):
    cmats.append(
        train_test1(features, labels, train, test))
return jug_sum(cmats)
cmatrix = nfoldcrossvalidation(features, labels)
output(cmatrix, dset, fset)
\end{python}

@@

\section{Development of Jug}
Jug is available under the MIT software license, which grants the user right to
use, copy, and modify the code almost at will. It is developed using the git
version control tool, with the project being hosted on the github
platform.\footnote{The jug project is available at
\url{https://www.github.com/luispedro/jug}.} An open mailing-list provides
support and reported bugs are generally fixed within hours or days.

Jug includes a full test suite with continuous integration, which are best
practices \citep{wilson2014best} which can improve productivity
\citep{vasilescu2015quality}. There are no known bugs.

@@

Several pipeline tools have been used (for a recent review, see the work of
\citet{leipzig2016a}). \citet{2007reproducible} proposed a system built on
top of Scons which shared superficially similar design. Scons, like Make, is a
build system. It supports spawning parallel jobs using multiple threads. In the
original use case, the tasks are delegated to other commands and the operating
system can take care of parallelism. If, however, the tasks are computationally
intensive in Python, contention for the global interpreter lock will limit the
amount of real parallelism.\footnote{In the most commonly used Python
interpreters, there is a lock that prevents more than one thread from
simultaneously executing interpreted code, although they can execute
non-interpreted code, as calling an external programme or certain external
libraries.} Ruffus is another Python-based solution which supports parallel
execution of pipelines \citep{goodstadt2010ruffus}. Using the Ruby programming
language, \citet{mishima2011} proposed Pwrake, which supports parallel
execution of bioinformatics workflows written (or accessible) in that language.
Snakemake~\citep{kster2012snakemakea} also improves over Make by providing more
complex rules and automatic interaction with a high performance compute
cluster. \citet{sadedin2012bpipe} proposed Bpipe, a tool based on a
domain-specific language around shell scripting. One large difference of Jug
compared to these tools is that Jug is tightly integrated with Python code.
Thus, while it lacks support for directly spawning processes like Make, it is
easier to call Python-based functions.

Also relevant is the \textsc{IncPy} implementation of a Python interpreter with
automatic persistent memoization~\cite{guo2010towards,guo2011using}
(unfortunately, no longer maintained). The advantages of that system apply to
Jug as well, with a few differences. Their system implements automatic
memoization, while, using Jug, the user needs to automatically annotate
functions to memoize using \texttt{TaskGenerator}.  This extra control can be
necessary to avoid a proliferation of intermediate results when these are very
large (often a very large intermediate output is not necessary, and only a
summary must be kept) at the cost of increased overhead for the programmer.
Additionally, Jug supports running tasks in parallel, a functionality that is
absent from \textsc{IncPy}.

Finally, joblib and memo \citep{moreno2016improving} use mechanisms similar to
Jug for in-process and in-memory memoization and parallelization.

@@

Jug focuses on the development of the computation as much as its communication.
In fact, when it comes to communication other tools might be better suited as
they combine written exposition with computer code. They can be used in
combination with jug as they do not often provide caching and distribution,
needed for large projects. While the simplest mode of jug use is to have a
single \textit{jugfile}, a jugfile can import the results of another. For
example, a first jugfile might perform all heavy-duty computation and be run on
a computing cluster. A second script could be embedded inside an executable
paper to generate tables and plots \citep{2011making}. This faster script could
be run as part of building the paper (while the longer script would take too
long).

Jug improves the pipeline development experience and makes the programming
researcher more productive and less error-prone. As many scientists now spend a
significant fraction of their time performing computational work
\citep{prabhu2011a,hannay2009how}, increasing their productivity in this task
can have significant effects.  Jug is based on Python, a widely used scientific
programming language, it can be used in a known environment without a high
learning curve.

Jug was not designed to compete with large scale frameworks such as Hadoop or
Spark, which scale better to very large projects. However, those frameworks
have much higher costs in terms of development time (code must be written
especially for the framework) and overhead. For scientists who want to quickly
adapt pre-existing code to run on a cluster or even a multi-core machine, jug
provides a better trade-off than those higher-powered alternatives in a
familiar programming language.
